The Evolution of Phenotypically Invariant Genetic Networks

### 2018-03-09 kalman_walk_paper.tex
a few more tweaks

% wright1932roles mentions possibility of equivalent combinations
% wright1935evolution looks at allele freq dynamics under stabilizing selection on a polygenic quantitative trait and discusses possibility for speciation this way
% barton1989divergence not available but likely following on from barton1986maintenance analyzing multiple optima in a similar model
% nei1983models simulates from simple stepwise mut model on one or two loci (without interactions) where incompat betw diff alleles increases with number of steps betw them
% sved1981twosex mathematically analyzes and (in another paper) simulates speciation in a model of male + female quantitative traits indep'ly evolving with attraction between matching traits
% barton2010natural has a good historical review of natural selection and speciation; then analyzes a two-niche levene model
% gavrilets2004fitness talks a lot about quasineutral paths in holey fitness landscapes
% phillips1996maintenance shows that subpop structure can stably maintain equivalent but incompat genotypes
% tulchinsky2014hybrid simulates a single promoter region with a bioenergentic model with directional and stabilizing selection, found incompat in both cases
% porter2002speciation simulated speciation under a simple model of a regulatory cascade
%%% Fisher's geom model: (incompat due to segregation variance)
% poon2000compensating also uses SSSWM with Fisher's model and derives some analytical results for load in small populations; early citation for this process
% slatkin1994segregation shows that segregation variance increases linearly with divergence if variation is controlled by intermediate freq alleles;
%   but if it is determined by rare alleles (at mut-sel balance), then it plateaus (at twice the within-pop value?)
% barton2001role looks at hybrids betw pops adapting to shifting optima in Fisher's geometric model : under fluctuating optima
%       hybrids will suffer because of cryptic segregating variation
% fraisse2016genetics uses SSWM; looks at stabilizing selection using Fisher's geometric model, showing it predicts lots of observations;
%        but mostly only if fitness is exp(-|x|^k) for k > 2
% chevin2014niche uses SSWM; looks at incompats after adaptation in Fisher's geom model and how it depends on dimension;
%        uses generalized Fisher's model with neutral directions; terms hybrid unfitness due to segregation variance "transgressive incompats";
%        finds rate of RI accumulation decreases with Ne (like Ne^(-m/2), with m the trait dimension?)
% simon2017coadapted follows up on the Fisher's geom model predictions, also comparing to some real data
% martin2014fishers shows that you can get Fisher's model quite generally if (a) the kryptotype dimension is large compared to the phenotype dimension,
%   and (b) mutational effects fall under the CLT
% rosas2010cryptic looks at segregation variance resulting in deviations from optimum phenotype including off of curvey ridges
%%%% General notes
% citation for segregation variance around a stationary optimum -> load is Poon & Otto 2000; rosas2010cryptic

It is not a new observation that there is often more than one way to do the same thing, 
or that speciation can be the result of (nearly) neutral processes.
The potential for speciation has been analyzed in models of 
traits under stabilizing selection determined additively by alleles at many loci 
\citep{wright1935evolution,barton1986maintenance,barton2001role},
in related fitness landscape models \citep{fraisse2016genetics},
and for pairs of traits that must match but whose value is unconstrained \citep{sved1981twosex}.
% Recent work has shown that Fisher's geometric model of stabilizing selection
% predicts many general patterns from the speciation literature \citep{fraisse2016genetics}.
It has also been shown that population structure
can allow long-term stable coexistence of incompatible genotypes encoding identical phenotypes \citep{phillips1996maintenance}.
However, previous simulations of system drift in regulatory sequences \citep{tulchinsky2014hybrid}
and a regulatory cascade \citep{porter2002speciation}
found rapid speciation under directional selection
but only equivocal support for speciation under models of purely neutral drift.
%Additionally, \citet{fierst} observed speciation due to an increase of segregation variance under balancing selection, in models of weak to moderate, but not strong epistasis.
The rate at which hybrid incompatibility accumulates due to genetic drift creating
segregation variance between isolated populations is fairly well understood
\citep{slatkin1994segregation,rosas2010cryptic,chevin2014niche},
but model assumptions can strongly affect predictions,
including whether variation is due to rare or common alleles \citep{slatkin1994segregation},
and the shape of the fitness landscape \citep{fraisse2016genetics}.
Our main aim is to provide a concrete framework 
that can provide natural predictions of these model parameters
across a general class of models.
Furthermore, tools from system theory allow analytical predictions to be made
for large populations with complex phenotypes
that would be inaccessible to population simulations.

@@

\paragraph{Haldane's rule}
This model naturally predicts
Haldane's rule, the observation that if only one hybrid sex is sterile or inviable 
it is likely the heterogametic sex (e.g., the male in XY sex determination systems) \citep{haldane1922sex, orr1997haldane}.
For example, consider an XY species with a two-gene network where the first gene resides on an autosome and the second gene on the X chromosome.
A male whose pair of haplotypes is 
$\left( \left[ \begin{smallmatrix} A_1 & A_2 \\ \cdot & \cdot \end{smallmatrix} \right], \left[ \begin{smallmatrix} A_1 & A_2 \\ X_1 & X_2 \end{smallmatrix} \right] \right)$
has phenotype determined by $A = \left[ \begin{smallmatrix} A_1 & A_2 \\ X_1 & X_2 \end{smallmatrix} \right]$, 
  if dosage compensation upregulates heterogametes by a factor of two relative to homogametes (as with \emph{Drosophila}),
while a female homozygous for the haplotype
$\left[\begin{smallmatrix} \bar A_1 & \bar A_2 \\ \bar X_1 & \bar X_2 \end{smallmatrix}\right]$, has phenotype determined by
$A = \left[\begin{smallmatrix} \bar A_1 & \bar A_2 \\ \bar X_1 & \bar X_2 \end{smallmatrix}\right]$.
An $F_1$ male offspring of these two will have its phenotype determined by
$\left[\begin{smallmatrix} (A_1 + \bar A_1)/2 & (A_2 + \bar A_2)/2 \\ \bar X_1 & \bar X_2 \end{smallmatrix}\right]$.
If both genes resided on the autosomes, this system would only be possible in an $F_2$ cross.
More generally, 
if the regulatory coefficients for a system are shared between the sex and one or more autosomal chromosomes, 
$F_1$ males are effectively equivalent to purely autosomal-system $F_2$ hybrids, and recall that $F_2$s are significantly less fit on average than $F_1$s (see Figure \ref{fig:osc_incompat}).
Although many alleles will be dominant if the phenotype--fitness relationship is convex,
the underlying mechanism does not depend on the \emph{dominance theory} \citet{turelli1995dominance} to explain Haldane's rule:
instead, it derives from the nature of segregation load.
% In this setting, Haldane's rule is mechanistically similar to \emph{hybrid breakdown}, 
% the observation that $F_2$ hybrids will often be less fit than $F_1$ hybrid crosses, in the early stages of speciation.
% This mechanism relies on the genetic variation of a system being shared across the sex chromosomes and the autosomes,
% and appears to be consistent with the ``dominance'' theory \citep{orr1997haldane, coyne1998evolutionary, turelli1995dominance} which is often invoked to explain Haldane's rule.
% \plr{Explain that the effects on fitness are dominant.}
%and makes several different predictions than the dominance theory as formulated by \citet{turelli1995dominance}.
%\plr{what are they?}
%\citet{turelli1995dominance} employ a population genetic model which strictly requires alleles causing hybrid incompatibility on the X chromosome not to exhibit additive effects, that is the \emph{dominance coefficient} (typically $h$ in classic population genetics) must not be equal to $\frac{1}{2}$. In their model, if alleles are strictly additive, that is neither dominant nor recessive, then male and female $F_1$ hybrids will drop in fitness at the exact same rates, and therefore defy Haldane's rule.

@@

%% REVIEW OF RELATED RESULTS
This fitness landscape is locally Gaussian with a rank-deficient covariance matrix.
Since we allow for substantial genetic variation within populations,
this model falls in the same class as \citet{lande1981models} and \citet{lande1983measurement},
which did not consider reproductive incompatibility.
A great deal of analytical work on the question of speciation
has been done under the assumption of a monomorphic population
whose trait is shifted by sequential fixation of alleles,
i.e., Fisher's ``geometric model'' \citep{fisher1930genetical,poon2000compensating}.
\citet{martin2014fishers} derived this model from a few general assumptions,
and used random matrix theory to calculate the distribution of fitness effects.
In particular, 
\citet{chevin2014niche} studied a general version with neutral directions,
and found that the rate of accumulation of reproductive isolation 
decreases with $N_e$ with a form that depends on trait dimension.
\citet{fraisse2016genetics} showed that most recognized empirical patterns in the speciation literature
could be explained by this model 
(although best if fitness took the form $\exp(-d^k)$
for some $k > 2$, where $d$ is distance to the optimum),
and \citet{simon2017coadapted} further compared predictions from the model to empirical data.
In the context of a stationary optimum, 
the primary contribution to hybrid unfitness is segregation variance,
i.e., greater phenotypic variance in $F_2$ hybrids than in parentals
due to drift having changed the genetic basis of the trait separately in the parental populations
(\citet{chevin2014niche} refers to these as ``transgressive incompatibilities'').
This turns out to be the main source of incompatibilities in the model we study as well,
even though in principle curvature of the optimal set also contributes
(as also seen in \citet{rosas2010cryptic}).
Our analysis relies on local approximations;
on longer time scales the appropriate model may share properties with
the ``holey'' landscapes of \citet{gavrilets2004fitness}.

As our goal here is to sketch out the rough implications of our main results,
keeping the assumptions clearly visible,
we provide often heuristic arguments,
rather than presenting calculations in full multivariate generality.

@@

\paragraph{Hybridization}
The means of two allopatric populations each of effective size $N_e$ separated for $T$ generations
will be a distance roughly of order $2\sigma_N \sqrt{T/N_e}$ apart along $\optx$.
(Consult figure \ref{fig:conceptual_fig} for a conceptual diagram.)
A population of $F_1$ hybrids has one haploid genome from each,
whose coefficients are averaged,
and so will have mean system coefficients at the midpoint between their means.
The distribution of $F_2$ hybrids will have mean at the average of the two populations,
but will have higher variance \citep{barton2001role,rosas2010cryptic,chevin2014niche}.
This ``segregation variance'' of $F_2$ hybrids 
can be shown to increase linearly with the square of the distance between
parental population means
under models of both simple and polygenic traits.
This is suggested by figure \ref{fig:conceptual_fig} and shown by \citet{slatkin1994segregation}
(also see Appendix \ref{apx:seg_cov} for a different derivation).
Concretely, we expect the population of $F_1$s to have variance $\sigma^2_S$ in the selected direction
(the same as within each parental population),
but the population of $F_2$ hybrids will have variance of order $\sigma^2_S + 4 \omega \sigma^2_N T/N_e$,
where $\omega$ is a factor that depends on the genetic basis of the coefficients
\citep[the ``segregation variance''][]{slatkin1994segregation}.
If the optimal set $\allS$ has dimension $q$,
using the polygenic model of appendix \ref{apx:seg_cov}, 
$\omega$ is proportional to the number of degrees of freedom: $\omega = (p-q)/8$.
If each trait is controlled by a single locus, as in figure \ref{fig:conceptual_fig},
the value is similar.

@@

\paragraph{Fisher's geometric model}
Substantial analytical work on speciation has been done using Fisher's geometric model, 
which can also predict Haldane's rule and the $F_2$ ``speciation clock'' 
\citep{barton2001role,fraisse2016genetics,simon2017coadapted}.
The model is similar to the quantitative genetics model we use 
to predict the rate of speciation, 
although work on Fisher's geometric model generally only models substitutions
between populations, i.e., neglecting within-species polymorphism.
Indeed, our model may provide an additional mechanistic context 
in which Fisher's geometric model is appropriate,
although the typical degree of pleiotropy
and stability of the $G$-matrix are unknown in practice.
Our observations argue for the importance of including neutral directions in these models,
which is not usually done (but see \citet{chevin2014niche}).

>>>>

% wright1932roles mentions possibility of equivalent combinations
% wright1935evolution looks at allele freq dynamics under stabilizing selection on a polygenic quantitative trait and discusses possibility for speciation this way
% barton1989divergence following on from barton1986maintenance analyzing multiple optima in a similar model, looking at RI due to shifts between equilibria
% nei1983models simulates from simple stepwise mut model on one or two loci (without interactions) where incompat betw diff alleles increases with number of steps betw them
% sved1981twosex mathematically analyzes and (in another paper) simulates speciation in a model of male + female quantitative traits indep'ly evolving with attraction between matching traits
% barton2010natural has a good historical review of natural selection and speciation; then analyzes a two-niche levene model
% gavrilets2004fitness talks a lot about quasineutral paths in holey fitness landscapes
% phillips1996maintenance shows that subpop structure can stably maintain equivalent but incompat genotypes
% tulchinsky2014hybrid simulates a single promoter region with a bioenergentic model with directional and stabilizing selection, found incompat in both cases
% porter2002speciation simulated speciation under a simple model of a regulatory cascade
%%% Fisher's geom model: (incompat due to segregation variance)
% poon2000compensating also uses SSSWM with Fisher's model and derives some analytical results for load in small populations; early citation for this process
% slatkin1994segregation shows that segregation variance increases linearly with divergence if variation is controlled by intermediate freq alleles;
%   but if it is determined by rare alleles (at mut-sel balance), then it plateaus (at twice the within-pop value?)
% barton2001role looks at hybrids betw pops adapting to shifting optima in Fisher's geometric model : under fluctuating optima
%       hybrids will suffer because of cryptic segregating variation
% fraisse2016genetics uses SSWM; looks at stabilizing selection using Fisher's geometric model, showing it predicts lots of observations;
%        but mostly only if fitness is exp(-|x|^k) for k > 2
% chevin2014niche uses SSWM; looks at incompats after adaptation in Fisher's geom model and how it depends on dimension;
%        uses generalized Fisher's model with neutral directions; terms hybrid unfitness due to segregation variance "transgressive incompats";
%        finds rate of RI accumulation decreases with Ne (like Ne^(-m/2), with m the trait dimension?)
% simon2017coadapted follows up on the Fisher's geom model predictions, also comparing to some real data
% martin2014fishers shows that you can get Fisher's model quite generally if (a) the kryptotype dimension is large compared to the phenotype dimension,
%   and (b) mutational effects fall under the CLT
% rosas2010cryptic looks at segregation variance resulting in deviations from optimum phenotype including off of curvey ridges
%%%% General notes
% citation for segregation variance around a stationary optimum -> load is Poon & Otto 2000; rosas2010cryptic

It is not a new observation that there is often more than one way to do the same thing, 
or that speciation can be the result of (nearly) neutral processes.
The potential for speciation has been analyzed in models of 
traits under stabilizing selection determined additively by alleles at many loci 
\citep{wright1935evolution,barton1986maintenance,barton1989divergence,barton2001role},
in related fitness landscape models \citep{fraisse2016genetics},
and for pairs of traits that must match but whose value is unconstrained \citep{sved1981twosex}.
% Recent work has shown that Fisher's geometric model of stabilizing selection
% predicts many general patterns from the speciation literature \citep{fraisse2016genetics}.
It has also been shown that population structure
can allow long-term stable coexistence of incompatible genotypes encoding identical phenotypes \citep{phillips1996maintenance}.
However, previous simulations of system drift in regulatory sequences \citep{tulchinsky2014hybrid}
and a regulatory cascade \citep{porter2002speciation}
found rapid speciation under directional selection
but only equivocal support for speciation under models of purely neutral drift.
%Additionally, \citet{fierst} observed speciation due to an increase of segregation variance under balancing selection, in models of weak to moderate, but not strong epistasis.
The rate at which hybrid incompatibility accumulates due to genetic drift creating
segregation variance between isolated populations is fairly well understood
\citep{slatkin1994segregation,rosas2010cryptic,chevin2014niche},
but model assumptions can strongly affect predictions,
including whether variation is due to rare or common alleles \citep{slatkin1994segregation},
and the shape of the fitness landscape \citep{fraisse2016genetics}.
Our main aim is to provide a concrete framework 
that can provide natural predictions of these model parameters
across a general class of models.
Furthermore, tools from system theory allow analytical predictions to be made
for large populations with complex phenotypes
that would be inaccessible to population simulations.

@@

\paragraph{Haldane's rule}
This model naturally predicts
Haldane's rule, the observation that if only one hybrid sex is sterile or inviable 
it is likely the heterogametic sex (e.g., the male in XY sex determination systems) \citep{haldane1922sex, orr1997haldane}.
For example, consider an XY species with a two-gene network where the first gene resides on an autosome and the second gene on the X chromosome.
A male whose pair of haplotypes is 
$\left( \left[ \begin{smallmatrix} A_1 & A_2 \\ \cdot & \cdot \end{smallmatrix} \right], \left[ \begin{smallmatrix} A_1 & A_2 \\ X_1 & X_2 \end{smallmatrix} \right] \right)$
has phenotype determined by $A = \left[ \begin{smallmatrix} A_1 & A_2 \\ X_1 & X_2 \end{smallmatrix} \right]$, 
  if dosage compensation upregulates heterogametes by a factor of two relative to homogametes (as with \emph{Drosophila}),
while a female homozygous for the haplotype
$\left[\begin{smallmatrix} \bar A_1 & \bar A_2 \\ \bar X_1 & \bar X_2 \end{smallmatrix}\right]$, has phenotype determined by
$A = \left[\begin{smallmatrix} \bar A_1 & \bar A_2 \\ \bar X_1 & \bar X_2 \end{smallmatrix}\right]$.
An $F_1$ male offspring of these two will have its phenotype determined by
$\left[\begin{smallmatrix} (A_1 + \bar A_1)/2 & (A_2 + \bar A_2)/2 \\ \bar X_1 & \bar X_2 \end{smallmatrix}\right]$.
If both genes resided on the autosomes, this system would only be possible in an $F_2$ cross.
More generally, 
if the regulatory coefficients for a system are shared between the sex and one or more autosomal chromosomes, 
$F_1$ males are effectively equivalent to purely autosomal-system $F_2$ hybrids, and recall that $F_2$s are significantly less fit on average than $F_1$s (see Figure \ref{fig:osc_incompat}).
Although many alleles will be dominant if the phenotype--fitness relationship is convex,
the underlying mechanism does not depend on the \emph{dominance theory} \citet{turelli1995dominance} to explain Haldane's rule:
instead, it derives from the nature of segregation variance.
% In this setting, Haldane's rule is mechanistically similar to \emph{hybrid breakdown}, 
% the observation that $F_2$ hybrids will often be less fit than $F_1$ hybrid crosses, in the early stages of speciation.
% This mechanism relies on the genetic variation of a system being shared across the sex chromosomes and the autosomes,
% and appears to be consistent with the ``dominance'' theory \citep{orr1997haldane, coyne1998evolutionary, turelli1995dominance} which is often invoked to explain Haldane's rule.
% \plr{Explain that the effects on fitness are dominant.}
%and makes several different predictions than the dominance theory as formulated by \citet{turelli1995dominance}.
%\plr{what are they?}
%\citet{turelli1995dominance} employ a population genetic model which strictly requires alleles causing hybrid incompatibility on the X chromosome not to exhibit additive effects, that is the \emph{dominance coefficient} (typically $h$ in classic population genetics) must not be equal to $\frac{1}{2}$. In their model, if alleles are strictly additive, that is neither dominant nor recessive, then male and female $F_1$ hybrids will drop in fitness at the exact same rates, and therefore defy Haldane's rule.

@@

%% REVIEW OF RELATED RESULTS
This fitness landscape is locally Gaussian with a rank-deficient covariance matrix.
Since we allow for substantial genetic variation within populations,
this model falls in the same class as \citet{lande1981models} and \citet{lande1983measurement},
which did not consider reproductive incompatibility.
Substantial work on speciation has been done under the assumption of a monomorphic population
whose trait is shifted by sequential fixation of alleles,
i.e., Fisher's ``geometric model'' \citep{fisher1930genetical,poon2000compensating}.
This work has included both stationary optima (like we study)
and moving optima (i.e., adaptation) \citep[e.g.,][]{barton2001role,chevin2014niche}.
\citet{martin2014fishers} derived this model from a few general assumptions,
and used random matrix theory to calculate the distribution of fitness effects.
\citet{chevin2014niche} studied a general version with neutral directions,
and found that the rate of accumulation of reproductive isolation 
decreases with $N_e$ with a form that depends on trait dimension.
\citet{fraisse2016genetics} showed that most recognized empirical patterns in the speciation literature
could be explained by this model 
(although best if fitness took the form $\exp(-d^k)$
for some $k > 2$, where $d$ is distance to the optimum),
and \citet{simon2017coadapted} further compared predictions from the model to empirical data.
In the context of a stationary optimum, 
the primary contribution to hybrid unfitness is segregation variance,
i.e., greater phenotypic variance in $F_2$ hybrids than in parentals
due to drift having changed the genetic basis of the trait separately in the parental populations
(\citet{chevin2014niche} refers to these as ``transgressive incompatibilities'').
This turns out to be the main source of incompatibilities in the model we study as well,
even though in principle curvature of the optimal set also contributes
(as also seen in \citet{rosas2010cryptic}).
Our analysis relies on local approximations;
on longer time scales the appropriate model may share properties with
the ``holey'' landscapes of \citet{gavrilets2004fitness}.

As our goal here is to sketch out the rough implications of our main results,
keeping the assumptions clearly visible,
we provide relatively rough arguments,
rather than presenting calculations in full multivariate generality.

@@

\paragraph{Hybridization}
The means of two allopatric populations each of effective size $N_e$ separated for $T$ generations
will be a distance roughly of order $2\sigma_N \sqrt{T/N_e}$ apart along $\optx$.
(Consult figure \ref{fig:conceptual_fig} for a conceptual diagram.)
A population of $F_1$ hybrids has one haploid genome from each,
whose coefficients are averaged,
and so will have mean system coefficients at the midpoint between their means.
The distribution of $F_2$ hybrids will have mean at the average of the two populations,
but will have higher variance \citep{wright1968evolution,barton2001role,rosas2010cryptic}.
This ``segregation variance'' of $F_2$ hybrids 
can be shown to increase linearly with the square of the distance between
parental population means
under models of both simple and polygenic traits.
This is suggested by figure \ref{fig:conceptual_fig} and shown by \citet{slatkin1994segregation}
(also see Appendix \ref{apx:seg_cov} for a different derivation).
Concretely, we expect the population of $F_1$s to have variance $\sigma^2_S$ in the selected direction
(the same as within each parental population),
but the population of $F_2$ hybrids will have variance $\sigma^2_S + 4 \omega \sigma^2_N T/N_e$,
where $\omega$ is a factor that depends on the genetic basis of the coefficients.
If the optimal set $\allS$ has dimension $q$,
using the polygenic model of appendix \ref{apx:seg_cov}, 
$\omega$ is proportional to the number of degrees of freedom: $\omega = (p-q)/8$.
If each trait is controlled by a single locus, as in figure \ref{fig:conceptual_fig},
the value is similar.

@@

\paragraph{Fisher's geometric model}
Substantial analytical work on speciation has been done using Fisher's geometric model, 
which can also predict Haldane's rule and the regular increase of incompatibility with genetic distance
\citep{barton2001role,fraisse2016genetics,simon2017coadapted}.
The model is similar to the quantitative genetics model we use 
to predict the rate of speciation, 
although work on Fisher's geometric model generally only models substitutions
between populations, neglecting within-species polymorphism.
Indeed, our model may provide an additional mechanistic context 
in which Fisher's geometric model is appropriate,
although the typical degree of pleiotropy
and stability of the $G$-matrix are unknown in practice.
Our observations argue for the importance of including neutral directions in these models,
which is not usually done (but see \citet{chevin2014niche}).
